Investigation of thermal reactionomes reveals differing responses to thermal

### 2014-10-01 doc/ApTranscriptome_manuscript.Rmd
updates to Intro with comments from SHC, NS, RD

Thermal tolerance is a key trait governing the distributions of species and underlies their future responses to climate change. Here, we characterize the thermal “reactionome” – the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient to understand the genetic basis of thermal tolerance. We characterized the reactionomes of two common ant species in eastern U.S. forests: the northern cold-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*. We found that at least 10% of genes were thermally responsive. The warm-adapted *A. carolinensis* displayed higher basal levels of gene expression, upregulated fewer heat shock proteins, and exhibited greater inertia of gene expression chang than did *A. picea*. This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability. *A. carolinensis* also shut down more genes in response to thermal extremes than *A. picea*. A subset of these genes were upregulated in *A. picea*, including gene ontology terms such as metabolism, cellular organization, glycoprotein synthesis and DNA repair. These differences between species suggest that the two closely related species employ different strategies for coping with temperature stress. Namely *A. picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas *A. carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress. More broadly, this study adds to a growing body of evidence that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.

@@

Successfully coping with progressive changes in climate will require both short-term physiological acclimation and longer-term adaptive changes. To better understand the relationship between these two timescales, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A. carolinensis*. We found surprisingly divergent transcriptomic responses in the two species. *A. picea* responded to high and low temperatures through upregulation of a larger proportion of its transcriptome with a more robust heat shock response. In contrast, *A. carolinensis* reacted to thermal stress to a greater extent through downregulation, reducing metabolic gene expression and requiring more extreme temperatures to trigger a protective response. These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely-encountered stressors, which may limit adaptive potential of northern species in the face of rapidly shifting climatic zones.

@@

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. Outside of candidate genes [@krebs1998], little is known about the genetic basis of this variation or how it has evolved. To address this, we need information on the extent to which geographic differentiation and selection has acted upon the diversity of genes, and their plasticity of expression, involved in thermal tolerance. This information is a critical for understanding the adaptive potential of species to future climates [@hoffmann2008] and the link between short-term physiological acclimation and longer-term adaptation to increased temperatures [@somero2010].

In recent years, whole organism gene expression approaches (e.g. transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes underlying thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014] and other traits. These RNAseq studies have used an ANOVA-type experimental or sampling design with only a few environmental levels. However, temperature and other environmental factors vary continuously in nature. As a result, many patterns of biologically-relevant plasticity are likely to be missed with ANOVA-style designs that compare only a few discrete levels of an environmental variable [@murren2014]. Similarly, gene expression studies that look at just a few environments (e.g. hot vs. cold) are likely to miss key differences among species that are due not just to whether it is hot, but rather how hot it is. What is needed is a reaction norm approach, which describes variation in the phenotype of a single genotype in response to a continuous environmental gradient [@gomulkiewicz1992]. Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values of either species or populations [@murren2014].

Regardless of the experimental design or treatment of temperature as a categorical variable with few levels, an important suite of issues often hampers many RNAseq studies that are designed to identify the genetic basis of environmental adaptation [@meyer2011; @teets2012; @barshis2013; @oneil2014]. Namely, lists of differentially-expressed genes derived from RNAseq results are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], may miss genes with small expression changes that have detectable phenotypic effects [@yan2002; @padovan2013], identify genes that are correlated, but not causative, of phenotypes [@fay2004] and may not be directly informative about organismal status [@feder2005]. Genome-wide association studies have faced a similar problem: phenotypic effects of a small number of statistically-significant genes are over-estimated while small effects of a large number of other causal genes are missed [@beavis1994; @larsson2013]. Our understanding of the genetic basis of traits, and our ability to predict evolutionary potential, are compromised by these issues.

Here, we present a novel method for RNAseq analysis: the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally-responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996]. Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006; @wardle2010].

To gain insight on the evolution of thermal adaptation, we compared the thermal reactionomes of these species with two specific predictions. First, species from more thermally-stressful environments will rely to a greater extent upon constitutive tolerance than plastically-induced resistance mechanisms [@barshis2013]. This pattern could be manifested through the number of thermally-responsive genes, or their sensitivity to changes in temperature. Finding that fewer genes are thermally responsive in the warmer-climate species *A. carolinensis* would support this prediction would support the former, while finding that more extreme temperatures are required to induce changes in gene expression in *A. carolinensis* would support the latter.

Second, we used the thermal reactionomes to examine the relationship between short-term physiological acclimation and longer-term adaptive changes to temperature. Specifically, through the process of genetic assimilation [@pigliucci2006], adaptive responses are hypothesized to involve the same molecular pathways as short-term acclimatory responses. Finding that the genes up-regulated in response to cold temperatures in *A. carolinensis* have increased basal expression in the colder-climate species *A. picea*, and vice versa, would be consistent with this prediction. While none of the above predictions depend on the functional annotation of the responsive genes [@pavlidis2012], we used gene set enrichment analysis to gain insight on the molecular toolkit involved in adaptation to thermal extremes. For example, in response to thermal extremes, we expect genes in the cellular stress response [@kultz2005] to be up-regulated, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated [@somero2011].

To test the above predictions for these two exemplar species, we collected whole organism gene expression data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina short read sequencing technology, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#options(scipen = 1, digits = 3)
load("ms_2014-09-25.RData")
library(data.table)
library(ggplot2)
library(reshape2)
library(pander)
library(RColorBrewer)
library(topGO)
```

>>>>

Thermal tolerance is a key trait governing the distributions of species and is likely useful in predicting their future responses to climate change. Here, we describe the thermal reactionome, the reaction norm for all genes in an organism's transcriptome measured across an experimental thermal gradient, to identify genes underlying thermal tolerance. We characterize the reactionomes of two common ant species in the eastern U.S: the northern cold-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*. We found that at least 10% of genes were thermally responsive. The warm-adapted *A. carolinensis* displayed higher basal levels of gene expression, upregulated fewer heat shock proteins, and exhibited greater inertia of gene expression change than did *A. picea* in response to warming. This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability. *Aphaenogaster carolinensis* also shut down more genes in response to thermal extremes than did *A. picea*, while a subset of these same genes were upregulated in *A. picea*, including gene ontology terms such as metabolism, cellular organization, glycoprotein synthesis and DNA repair. These differences indicate that the two closely related species employ different strategies for coping with temperature stress. The northern *A. picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas the southern *A. carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress. These results are in line with expectations that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.

@@

Coping with progressive changes in climate requires both short-term physiological acclimation and long-term adaptive changes. To better understand the relationship between these two timescales, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A. carolinensis*. Surprisingly, we found divergent transcriptomic responses in these species. *Aphaenogaster picea* responded to high and low temperatures through upregulation of a larger proportion of its transcriptome with a robust heat shock response. In contrast, *A. carolinensis* required more extreme temperatures to trigger a protective response and downregulated genes related to metabolism. These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely encountered stressors, which may limit adaptive potential of northern species in the face of rapidly shifting climatices.

@@

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Many species, especially small insects, live at temperatures close to their thermal limits and are at risk from current increases in temperature (sunday, kingsolve, deutch, hoffman). Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activiation of physiological response mechanisms to prevent and/or repair damage (...refs...). Thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. Outside of candidate genes [@krebs1998], little is known about the molecular basis of such variation or the link between short-term physiological acclimation and longer-term adaptation to novel temperature regimes. To address this gap of knowlege, we need information on the diversity of genes, and their plasticity of expression, involved in thermal tolerance [@hoffman2008; @somero2011].

In recent years, whole organism gene expression approaches (e.g. transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes involved in thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014] and other traits. Such studies typically use an ANOVA-type experimental or sampling design, with only a few environmental levels, and often find only a few dozen to hundred genes with differential expression in different thermal regimes. However, temperature and other environmental factors vary continuously in nature. As a result, many patterns of biologically-relevant plasticity are likely to be missed in studies that compare only a few levels of an environmental variable [@murren2014]. Similarly, species comparisons look at just a few environments (e.g. hot vs. cold) are likely to miss key differences among species that are due not just to whether it is hot, but rather how hot it is. Such differences in degree are better characterized with a reaction norm approach, which describes variation in the phenotype of a single genotype in response to a continuous environmental gradient [@gomulkiewicz1992]. Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values of either species or populations [@murren2014].

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally-responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster picea* are found ... while *A. carolinensis* occupy habitats with ... . Experimental warming experiments indicate that *Aphaenogaster* abundance and competitive ability may be significantly impacted by temperature increases (...refs...).

To gain insight on the evolution of thermal adaptation, we compared the thermal reactionomes of these species to test two specific predictions. First, we predicted that species from more thermally-stressful environments will rely to a greater extent than species from cooler environments upon constitutive tolerance than plastically-induced resistance mechanisms [@barshis2013]. This pattern could be manifested through the number of thermally-responsive genes, or their sensitivity to changes in temperature. Finding that fewer genes are thermally responsive in the warmer-climate species *A. carolinensis* would support this prediction would support the former, while finding that more extreme temperatures are required to induce changes in gene expression in *A. carolinensis* would support the latter. Second, we predicted that long-term adaptive changes to temperatures would involve the same genes that are involved in short-term physiological acclimation (e.g. genetic assimilation [@pigliucci2006]). Finding that the genes up-regulated in response to cold temperatures in *A. carolinensis* have increased basal expression in the colder-climate species *A. picea*, and vice versa, would be consistent with this prediction.

While neither of these predictions depend on the functional annotation of the responsive genes [@pavlidis2012], we used gene set enrichment analysis to gain insight on the molecular toolkit involved in adaptation to thermal extremes. For example, in response to thermal extremes, we expect genes in the cellular stress response [@kultz2005] to be up-regulated, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated [@somero2011].

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#options(scipen = 1, digits = 3)
load("ms_2014-09-26.RData")
library(data.table)
library(ggplot2)
library(reshape2)
library(pander)
library(RColorBrewer)
library(topGO)
```
